---
title: "Set-up Instructions"
output: 
  html_document:
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(png)
```

I ask that you please complete at least the R and R Studio installation before our session. If you feel comfortable, you can go through the installation of packages and setting up the packages, but we will probably discuss this in class. Please [contact me](https://info.library.okstate.edu/clarke-iakovakis) if you have questions or problems with any of this.

# Download R & R Studio

To download R, go to <https://www.r-project.org/>. Click on CRAN (Comprehensive R Archive Network) under Download, and scroll down to your country. Select the download link corresponding to the city that is geographically closest to you.

Go to <https://www.rstudio.com/products/RStudio/#Desktop> to download the RStudio desktop software. RStudio is a user interface for working with R. It is called an Integrated Development Environment (IDE) and acts as a sort of wrapper around the R language. You can use R without RStudio, but it's much more limiting. RStudio makes it easier to import datasets, create and write scripts, and has an autocomplete activated for functions and variables you've already assigned. RStudio makes using R much more effective, and is also free and open source.

# Install R packages

Open R Studio and go to **File > New File > R Script**. 

Copy the first line of code below and paste it into the blank R script. Then click the **Run** button in the upper right corner of that window.

```{r install, eval=FALSE,cache=TRUE}
install.packages("rorcid")
library(rorcid)
```

It will print some lines of code and make take a few seconds. If successful, it will tell you the packaged was successfully unpacked. Now run the next line, `library(rorcid)`. As long as it doesn't say "there is no package called 'rorcid', you should be good.

Do the same thing with the remaining packages:

```{r install2, eval=FALSE,cache=TRUE}
install.packages("tidyverse")
library(tidyverse)

install.packages("anytime")
library(anytime)

install.packages("httpuv")
library(httpuv)

install.packages("rcrossref")
library(rcrossref)

install.packages("usethis")
library(usethis)

install.packages("janitor")
library(janitor)

install.packages("roadoi")
library(roadoi)
```

Now in R Studio, click **Session > Restart R.**

# Set up `rorcid`

Next, you need to authenticate with an ORCID API Key. According to the [ORCID API tutorial](https://members.orcid.org/api/tutorial/read-orcid-records), anyone can receive a key to access the public API. 

Make sure your R script is still open. If it's not, click **File > New File > R Script.**

Copy the below two lines of code and paste them in your new R Script. Execute them by clicking the Run button in the top right hand corner.

```{r install2x, eval=FALSE,cache=TRUE}
library(rorcid)
orcid_auth()
```

You should see a message stating: `no ORCID token found; attempting OAuth authentication` and a window will open in your default internet browser. Log-in to your orcid account. You will be asked to give `rorcid` authorization to access your ORCID Record for the purposes of getting your ORCID iD. Click "Authorize."

If successful, the browser window will state: "Authentication complete. Please close this page and return to R." Return to R Studio and you should see in your R console the word **Bearer**, followed by a long string of letters and numbers. These letters and numbers are your API key. At this point, this should be cached locally in your working directory. 

Highlight and copy the API key (the letters and numbers only--exclude the word "Bearer" and the space). Paste it into the quotation marks for the code below. Be sure to keep the quotation marks there. Don't run this line of code. Instead, just copy it to the clipboard.

```{r install2x1, eval=FALSE,cache=TRUE}
ORCID_TOKEN="copy and paste your token here"
```

Now paste the below text to your R Script and run it:

```{r install2x2, eval=FALSE,cache=TRUE}
library(usethis)
usethis::edit_r_environ()
```

A new window will open in R Studio. Type `ORCID_TOKEN="my-token"`, replacing `my-token` with the API key. **Then press enter to create a new line, and leave it blank.** It will look like something like this (below is a fake token):

```{r addkey, eval=FALSE}
ORCID_TOKEN="4bed1e13-7792-4129-9f07-aaf7b88ba88f"
```

Press <kbd>Ctrl</kbd> + <kbd>S</kbd> (Mac: <kbd>Cmd</kbd> + <kbd>S</kbd> to save the API key to your R environment and close the tab In R Studio, click Session > Restart R. Your token should now be saved to your R environment. You can confirm this by calling `orcid_auth()`, and it will print the token.



# Set up `rcrossref`

As described in the documentation, the Crossref team encourages requests with appropriate contact information, and will forward you to a dedicated API cluster for improved performance when you share your email with them. Learn more [here](https://github.com/CrossRef/rest-api-doc#good-manners--more-reliable-service). To do so, first open your R environment using the `edit_r_environ()` function from the `usethis` package.

```{r editenviron, eval=FALSE}
library(usethis)
usethis::edit_r_environ()
```

A new window will open in R Studio. Type `crossref_email=name@example.com`, replacing the example with your own email address. Then press enter to create a new line, and leave it blank. Press <kbd>Ctrl</kbd> + <kbd>S</kbd> (Mac: <kbd>Cmd</kbd> + <kbd>S</kbd> to save the API key to your R environment and close the tab Your email address will now be shared with Crossref. 


# Set up `roadoi`

As with `rcrossref`, your API calls to Unpaywall must include a valid email address where you can be reached in order to keep the service open and free for everyone. 

Run this line of code, replacing the example with your email address:

```{r roadoi1, eval=FALSE,cache=TRUE}
options(roadoi_email = "name@example.com")
```

Your email address will now be shared with Unpaywall.