---
title: "Work"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy=TRUE, cache = TRUE)
library(kableExtra)
library(knitr)
library(tidyverse)
library(tibble)
library(prettydoc)
library(listviewer)
library(janitor)
library(rorcid)
```

```{r myorcid, eval=TRUE, echo=FALSE}
my_orcid_id <- "0000-0002-9260-8456"
my_works <- rorcid::works(my_orcid_id) %>%
  janitor::clean_names() %>%
  dplyr::mutate(created_date_value = anytime::anydate(created_date_value/1000))
my_works_ex <- my_works %>% 
  dplyr::filter(!purrr::map_lgl(external_ids_external_id, purrr::is_empty)) %>% 
  tidyr::unnest(external_ids_external_id, .drop = TRUE) %>% 
  dplyr::bind_rows(my_works %>% 
                     dplyr::filter(map_lgl(external_ids_external_id,
                                           is_empty)) %>%
                     dplyr::select(-external_ids_external_id))

```












## [Building your Research Profile](https://doi.org/10.17605/osf.io/2w6pj)
Slides and recording on building your research profile, including a review of academic social networks, harvesting your publications, copyright, and sharing scholarship.




## [Open Access Outreach Materials: Slides & Handout](https://doi.org/10.6084/m9.figshare.7849160)

A set of resources for open access outreach. Includes 5 slideshows and a handout on the following: 

- Copyright in Higher Education
- Open Access: Increase the Visibility of your Scholarship
- Open Access: Identifying Quality Journals & Avoiding Predatory Publishers
- Open Educational Resources: An Introduction
- Open Data: Enhancing Preservation, Reproducibility, and Innovation
- Strategies for Accessing Research After Graduation

