---
title: "Seeking help"
output: html_document
---

The following lesson is derived from work that is copyright &copy; Software Carpentry at (http://swcarpentry.github.io/r-novice-gapminder/03-seeking-help/index.html) and made available under a [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/) license. The use of this lesson in the present course is not formally endorsed by Software Carpentry.

# Reading Help files

# R, and every package, provide help files for functions. The general syntax to search for help on any function, "function_name", from a specific function that is in a package loaded into your namespace (your interactive R session):

?function_name
help(function_name)


# This will load up a help page in RStudio (or as plain text in R by itself).

Each help page is broken down into sections:

 - Description: An extended description of what the function does.
 - Usage: The arguments of the function and their default values.
 - Arguments: An explanation of the data each argument is expecting.
 - Details: Any important details to be aware of.
 - Value: The data the function returns.
 - See Also: Any related functions you might find useful.
 - Examples: Some examples for how to use the function.

Different functions might have different sections, but these are the main ones you should be aware of.

> ## Tip: Running Examples
>
> From within the function help page, you can highlight code in the 
> Examples and hit <kbd>Ctrl</kbd>+<kbd>Return</kbd> to run it in 
> RStudio console. This is gives you a quick way to get a feel for
> how a function works.
> ## Tip: Reading help files
>
> One of the most daunting aspects of R is the large number of functions
> available. It would be prohibitive, if not impossible to remember the
> correct usage for every function you use. Luckily, the help files
> mean you don't have to!
{: .callout}
## Special Operators

To seek help on special operators, use quotes:

```{r, eval=FALSE}
?"<-"
```

## Getting help on packages

Many packages come with "vignettes": tutorials and extended example documentation.
Without any arguments, `vignette()` will list all vignettes for all installed packages;
`vignette(package="package-name")` will list all available vignettes for
`package-name`, and `vignette("vignette-name")` will open the specified vignette.

If a package doesn't have any vignettes, you can usually find help by typing
`help("package-name")`.

## When you kind of remember the function

If you're not sure what package a function is in, or how it's specifically spelled you can do a fuzzy search:

```{r, eval=FALSE}
??function_name
```

## When you have no idea where to begin

If you don't know what function or package you need to use
[CRAN Task Views](http://cran.at.r-project.org/web/views)
is a specially maintained list of packages grouped into
fields. This can be a good starting point.

## When your code doesn't work: seeking help from your peers

If you're having trouble using a function, 9 times out of 10,
the answers you are seeking have already been answered on
[Stack Overflow](http://stackoverflow.com/). You can search using
the `[r]` tag.

If you can't find the answer, there are a few useful functions to
help you ask a question from your peers:

```{r, eval=FALSE}
?dput
```

Will dump the data you're working with into a format so that it can
be copy and pasted by anyone else into their R session.

```{r}
sessionInfo()
```

Will print out your current version of R, as well as any packages you
have loaded. This can be useful for others to help reproduce and debug
your issue.


## Other ports of call

* [Quick R](http://www.statmethods.net/)
* [RStudio cheat sheets](http://www.rstudio.com/resources/cheatsheets/)
* [Cookbook for R](http://www.cookbook-r.com/)
